Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway

### 2021-04-14 analysis/paper/paper.Rmd
Maps done, some additional writing

```{r get-data, eval = TRUE, message = FALSE, warning = FALSE}
assemblages <- read.csv(here::here('analysis/data/raw_data/assemblage_data.csv'))
site_pts <- st_read(here::here('analysis/data/raw_data/sites.gpkg'))
countries <- ne_download(scale = 10, type = 'countries', category = 'cultural',
                         returnclass = "sf")
```

@@

The `r nrow(sites)` coastal sites chosen for analysis here have a relatively limited geographical distribution in south-eastern Norway. The sites were excavated as part of four larger excavation projects that all took place within the last 15 years [@solheim2013b; @jaksland2014a; @melvold2014b; @reitan2014l; @solheim2017b]. The sites included here consists of all Mesolithic sites excavated in conjunction with the projects with assemblages holding more than 100 artefacts. The institution responsible for all of these excavations was the Museum of Cultural History in Oslo. This has led to a considerable overlap in the archaeological personnel involved, and comparable excavation practices across the excavations. Furthermore, with these projects, major efforts were made to standardise how lithic artefacts were to be classified at the Museum of Cultural History. As a result, this should reduce the amount of artificial patterning in the data incurred by discrepancies in the employed systems for categorisation [@clark2006; @dibble2017]. In this setting, for example, bias could potentially follow from the fact that two of the projects have sites with relatively contemporaneous dates [@solheim2013b; @jaksland2014]. Any project-dependent classification practice could as a consequence lead to an exaggeration of chronological differences between the assemblages. While this is difficult to fully account for, I do believe that the relative contemporaneity of the excavation projects, as well as the overlap in excavation and classification practices should minimise the above-mentioned effects, making the data-set a good candidate for exploratory data analysis.

```{r overview}
# Create a bounding box around site points to limit the overview map
sitbbox <- st_bbox(site_pts)
sitbbox[1:2] <- sitbbox[1:2] - 500000
sitbbox[3:4] <- sitbbox[3:4] + 500000
boundingpoly <- st_as_sf(st_as_sfc(sitbbox))

@@

# Assemble overview map
overview <-
  ggplot() +
  geom_sf(data = count_reproj, fill = "grey", colour = NA) +
  geom_sf(data = boundingpoly2,
          fill = NA, colour = "black", size = 1) +
  coord_sf(xlim = c(sitbbox[1], sitbbox[3]), ylim = c(sitbbox[2], sitbbox[4]),
         expand = FALSE) +
  theme_bw() + theme(axis.title=element_blank(),
                     axis.text.y = element_blank(),
                     axis.text.x = element_blank(),
                     rect = element_rect(),
                     axis.ticks = element_blank(),
                     panel.grid.major = element_blank())

@@

# Assemble site map. anc defined first is used for placing the scalebar,
# which required some tweaking to get right.
anc <- as.numeric(c(sitbbox2$ymin, sitbbox2$xmax))
sitemap <-
  ggplot() +
  geom_sf(data = count_reproj2, fill = "grey", colour = NA) +
  geom_sf(data = sitejoin, aes(fill = excavation_project),
          size = 2.25, shape=21, 
          colour = "black") +
  scale_fill_manual(labels = c("E18 Bommestad-Sky", "E18 Brunlanes",
                               "E18 Rugtvedt-Dørdal", "Vestfoldbaneprosjektet"),
                    values = c("springgreen4", "royalblue2",
                                "gold1", "firebrick3")) +
  ggsn::scalebar(data = site_pts, dist = 10, dist_unit = "km", 
                 transform = FALSE, st.size = 4, height = 0.05, 
                 border.size = 0.05, st.dist = 0.05,
                 anchor = c(x = anc[2] - 8000, y = anc[1]) + 4000) +
  coord_sf(xlim = c(sitbbox2[1], sitbbox2[3]), 
           ylim = c(sitbbox2[2], sitbbox2[4]),
         expand = FALSE) +
  labs(fill = "Excavation project:") +
 theme_bw() + theme(axis.title=element_blank(),
                     axis.text.y = element_blank(),
                     axis.text.x = element_blank(),
                     rect = element_rect(),
                     axis.ticks = element_blank(),
                     panel.grid.major = element_blank(),
                    legend.position = "bottom")

yrs <- seq(11500, 5000, -500)
temp <- ggplot(sites, aes(x = yrs, y = reorder(site, mean_date))) +
  geom_segment(aes(x = reported_start_bp, xend = reported_end_bp,
                   yend = reorder(site, mean_date),
                   colour = excavation_project), size = 1.7) +
  xlab("Cal. BP") +
  scale_x_reverse() +
  scale_color_manual(values = c("springgreen4", "royalblue2",
                                "gold1", "firebrick3")) +
  theme_classic() +
  theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        legend.position = "none")
```

A defining characteristic of the Norwegian Mesolithic is that a clear majority of the known sites are located in coastal areas. Furthermore, these coastal sites, including those treated here, appear to predominantly have been located on or close to the contemporary shoreline when they were in use [@breivik2018; @solheim2020]. In south-eastern Norway, this pattern is combined with a continuous regression of the shoreline, following from isostatic rebound [e.g. @sørensen1979; @romundset2018]. The fairly rapid shoreline displacement means that the sites tend not to have retained their strategic or ecologically beneficial shore-bound location for long periods of time [cf. @perreault2019, 47]. Consequently, the shore-bound settlement, combined with the rapid shoreline displacement has resulted in a relatively high degree of spatial separation of cumulative palimpsests, to follow the terminology of @bailey2007, while the shoreline displacement combined with shore-bound settlement allows for a relatively good chronological control of these accumulation events. In other parts of the world, a higher degree of spatial distribution means that while the physical separation of material can help delineate discrete events this typically comes at the cost of loosing temporal resolution, as any stratigraphic relationship between the events is lost [@bailey2007]. However, as the rate of isostatic rebound has varied throughout the Mesolithic in the region, and local topography and bathymetry will have impacted how rapidly a site lost its shore-bound location, this effect is not evenly distributed in time and space. In the earliest part of the Mesolithic, the displacement rate within the study area would have been around as much as 8.8 cm/year, falling to around 0.5 cm/year in the Late Mesolithic [@sørensen2014]. Thus, while relative sea-level change appears to have reduced the degree of mixing that has occurred in the assemblages, this could vary depending on when and where they were in use, in turn potentially reducing the degree to which their composition can be directly compared - a point that is returned to in the discussion of the results.

@@

The first part of the analysis involves employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports. The purpose of this exercise is to evaluate the degree to which the composition of the assemblages align with patterns that have been suggested by earlier application of more informally driven methods. The expectation is that chronological trends will be the main driver of this part of the analysis. While some artefact categories from the original reports are collapsed in an attempt to reduce the amount of subjective discretion that can be expected to have impacted the classification of the individual assemblages, this will largely involve an analysis that assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviorally meaningful.

@@

```{r co_data}
# Perform correspondence analysis on the data. 'mean_age' is meant for colouring
# so is excluded here
calithics <- CA(select(lithics, -mean_date, -artefact_count, -flint_count), graph = FALSE)

@@

The column plot (Figure \@ref(fig:cor)B) reveals that the earliest sites are characterised by the flint artefact categories microburins, projectiles, as well as macro tools and associated debitage. It is also interesting that these sites to larger extent are characerised by core fragments, both in flint and non-flint materials, while the cores themselves appear to be a trait more defining of assemblages with later dates. The non-flint material on the earliest sites appears to be centered around the production of projectiles, as both the projectiles themselves and non-flint blades are important constituents of the assemblages at these sites.

The first dimension, which is pulling some of the later sites towards the right of the plot, is mainly defined by macro tools and associated debitage in non-flint materials that are negatively correlated with more flint dominated assemblages and non-flint projectiles. While the outer end of the first dimensions is dominated by later Mesolithic sites associated with axe production in non-flint materials, the later sites occur along the entire dimension, indicating that while these axe production sites are a feature of the later Mesolithic, there is marked variation among these later sites. Although the sample size is quite strained and the discussion of finer chronological points might not be warranted, the first dimension does appear to be of of less importance for the absolute latest sites, as indicated by their location to the left of the plot. This could indicate that specialised axe production sites disappear towards the end of the Mesolithic, a notion that would be in line with previous research.

@@

It has, however, been suggested that access to raw-materials will be a more important decider for assemblage composition than mobility patterns [@andrefsky1994], and that variable workability of these materials will impact factors such as the relative frequency of tools to debitage in the assemblages [@manninen2014]. Raw-material accessibility in south-eastern Norway is arguably understudied,. In addition to attempts at correcting for this in the analysis conducted here (see below), the limited geographical distribution of the sites should alleviate the issue somewhat, as the overall availability of raw materials could be expected to be more comparable through time than if a wider region was under consideration.

@@

lithics2 <- sites %>% 
  select(-n, -beach_flint_flint_nodule, -hammerstone, -excavation_project) %>% 
  # The reason for not using .keep below is for the option to retain artefact_count
  mutate(chip_freq =  sum(chip, chip_other) / artefact_count,
         core_freq = sum(core, core_other) / artefact_count,
         primary_freq = sum(core, core_fragment, blade, 
                     microblade, chip, fragment, flake, flake_other, 
                     fragment_other, chip_other, blade_other,  microblade_other, 
                     core_other, core_fragment_other) / artefact_count,
         secondary_freq =  sum(axe_debitage, axe, chisel, 
                microlith, tanged_point, transverse_point, single_edged_point,
                hognipen_point, projectile_undefined, microburin, scraper, 
                knife, burin, drill, blade_retouch, microblade_retouch, 
                chip_retouch, fragment_retouch, flake_retouch, axe_other,
                axe_debitage_other, chisel_other, club_hatchet_other, 
                grinding_slab_stone, projectile_point_other, knife_other,
                scraper_other, drill_other, microblade_retouch_other, 
                blade_retouch_other, fragment_retouch_other,
                flake_retouch_other) / artefact_count,
         lithic_dens = artefact_count / excavated_m3) %>%
  select(-primary_lithics, -secondary_lithics, -excavated_m3, -artefact_count,
         -flint_count) %>% 
  column_to_rownames("site")

flint2 <- sites %>% 
  select(-n, -beach_flint_flint_nodule, -hammerstone, -excavation_project) %>% 
  select(!ends_with("other")) %>% 
  # The reason for not using .keep below is for the option to retain artefact_count
  mutate(chip_freq =  sum(chip) / artefact_count,
         core_freq = sum(core) / artefact_count,
         primary_freq = sum(core, core_fragment, blade, 
                     microblade, chip, fragment, flake) / artefact_count,
         secondary_freq =  sum(axe_debitage, axe, chisel, 
                microlith, tanged_point, transverse_point, single_edged_point,
                hognipen_point, projectile_undefined, microburin, scraper, 
                knife, burin, drill, blade_retouch, microblade_retouch, 
                chip_retouch, fragment_retouch, flake_retouch) / artefact_count,
         flint_dens = flint_count / excavated_m3) %>%
  select(-primary_flint, -secondary_flint, -excavated_m3, -artefact_count,
         -flint_count) %>% 
  column_to_rownames("site")

@@

```{r wabi, fig.cap = "Whole-assemblage behavioural index", fig.height = 5, fig.width = 10}
w1 <- ggplot(lithics2, aes(lithic_dens, secondary_freq * 100, 
                     fill = mean_date / 1000)) + 
  geom_point(shape = 21, size = 3) +
  geom_smooth(method = "lm", col = "black", se = FALSE) +
  labs(x = "Volumetric density of lithics (log10)", 
       y = "Relative frequency of secondarily worked lithics (log10)") +
  scale_fill_gradient2(name = "ka BP",  low="yellow", mid="orange", 
                        high="red", midpoint = centdate) +
  scale_y_log10() +
  scale_x_log10() +
  theme_classic()

w2 <- ggplot(flint2, aes(flint_dens, secondary_freq * 100, 
                     fill = mean_date / 1000)) + 
  geom_point(shape = 21, size = 3) +
  geom_smooth(method = "lm", col = "black", se = FALSE) +
  labs(x = "Volumetric density of flint (log10)", 
       y = "Relative frequency of secondarily worked flint (log10)") +
  scale_fill_gradient2(name = "ka BP",  low="yellow", mid="orange", 
                        high="red", midpoint = centdate) +
  scale_y_log10() +
  scale_x_log10() +
  theme_classic()

@@

Figure \@ref(fig:pca-analysis) displays a principle components analysis using variations of the continuous measures for degree of mobility as operationalised by @Bicho2020. While the investigation performed by @Bicho2020 indicates that the relative frequency of chips, cores, and blanks might be more sensitive to mobility patterns than the volumetric density of lithics, and the frequency of retouched artefacts, these last dimensions are capturing the most variation in the dataset presented here. If frequency of secondarily worked artefacts is taken as a proxy for mobility, these findings would consequently be in line with previous research into the Mesolithic of Norway, indicating that earlier sites are associated with higher degree of mobility than sites from later phases. @Bicho2020 also found that WABI, as introduced by, gave a reasonable estimation of mobility in their data. The same is evident here where fig indicates a

>>>>

```{r get-data, eval = TRUE, results = 'hide'}
assemblages <- read.csv(here::here('analysis/data/raw_data/assemblage_data.csv'))
site_pts <- st_read(here::here('analysis/data/raw_data/sites.gpkg'))
countries <- ne_download(scale = 10, type = 'countries', category = 'cultural',
                         returnclass = "sf")
```

@@

The `r nrow(sites)` coastal sites chosen for analysis here have a relatively limited geographical distribution in south-eastern Norway \ref@(fig:dist)A]. The sites were excavated as part of four larger excavation projects that all took place within the last 15 years [@solheim2013b; @jaksland2014a; @melvold2014b; @reitan2014l; @solheim2017b]. The sites included here consists of all Mesolithic sites excavated in conjunction with the projects with assemblages holding more than 100 artefacts. The institution responsible for all of these excavations was the Museum of Cultural History in Oslo. This has led to a considerable overlap in the archaeological personnel involved, and comparable excavation practices across the excavations. Furthermore, with these projects, major efforts were made to standardise how lithic artefacts were to be classified at the Museum of Cultural History. As a result, this should reduce the amount of artificial patterning in the data incurred by discrepancies in the employed systems for categorisation [@clark2006; @dibble2017]. In this setting, for example, bias could potentially follow from the fact that two of the projects have sites with relatively contemporaneous dates [@solheim2013b; @jaksland2014, see also \ref@(fig:dist)B]. Any project-dependent classification practice could as a consequence lead to an exaggeration of chronological differences between the assemblages. While this is difficult to fully account for, I do believe that the relative contemporaneity of the excavation projects, as well as the overlap in excavation and classification practices should minimise the above-mentioned effects, making the data-set a good candidate for exploratory data analysis.

```{r dist, fig.cap = "A) Spatial and B) temporal distribution of the sites chosen for analysis. The date ranges follow those given in the original excavation reports", fig.height = 10, fig.width = 12}
# Create a bounding box around site points to limit the overview map
sitbbox <- st_bbox(site_pts)
sitbbox[1:2] <- sitbbox[1:2] - 600000
sitbbox[3:4] <- sitbbox[3:4] + 600000
boundingpoly <- st_as_sf(st_as_sfc(sitbbox))

@@

# Assemble overview map
overview <-
  ggplot() +
  geom_sf(data = count_reproj, fill = "grey", colour = NA) +
  geom_sf(data = boundingpoly2,
          fill = NA, colour = "black", size = 1) +
  coord_sf(xlim = c(sitbbox[1], sitbbox[3]), ylim = c(sitbbox[2], sitbbox[4]),
         expand = FALSE) +
  labs(title = "A") +
  theme_bw() + theme(axis.title=element_blank(),
                     axis.text.y = element_blank(),
                     axis.text.x = element_blank(),
                     rect = element_rect(),
                     axis.ticks = element_blank(),
                     panel.grid.major = element_blank())

@@

# Assemble site map. anc defined first is used for placing the scalebar,
# which required some tweaking to get right.
anc <- as.numeric(c(sitbbox2$ymin, sitbbox2$xmax))
sitemap <-
  ggplot() +
  geom_sf(data = count_reproj2, fill = "grey", colour = NA) +
  geom_sf(data = sitejoin, aes(fill = excavation_project),
          size = 2.25, shape=21, 
          colour = "black") +
  labs(fill = "Excavation project") +
  scale_fill_manual(labels = c("E18 Bommestad-Sky", "E18 Brunlanes",
                               "E18 Rugtvedt-Dørdal", "Vestfoldbaneprosjektet"),
                    values = c("springgreen4", "royalblue2",
                                "gold1", "firebrick3")) +
  ggsn::scalebar(data = site_pts, dist = 10, dist_unit = "km", 
                 transform = FALSE, st.size = 4, height = 0.05, 
                 border.size = 0.05, st.dist = 0.07,
                 anchor = c(x = anc[2] - 8000, y = anc[1]) + 4000) +
  coord_sf(xlim = c(sitbbox2[1], sitbbox2[3]), 
           ylim = c(sitbbox2[2], sitbbox2[4]),
         expand = FALSE) +
 theme_bw() + theme(axis.title=element_blank(),
                     axis.text.y = element_blank(),
                     axis.text.x = element_blank(),
                     rect = element_rect(),
                     axis.ticks = element_blank(),
                     panel.grid.major = element_blank(),
                     legend.position = "none") +
  guides(fill = guide_legend(ncol = 2,
                             title.position = "top", title.hjust = 0.5))

yrs <- seq(11500, 5000, -500)
temp_dist <- ggplot(sites, aes(x = yrs, y = reorder(site, mean_date))) +
  geom_segment(aes(x = reported_start_bp, xend = reported_end_bp,
                   yend = reorder(site, mean_date),
                   colour = excavation_project), size = 1.7) +
  xlab("Cal. BP") +
  scale_x_reverse() +
  scale_colour_manual(labels = c("E18 Bommestad-Sky", "E18 Brunlanes",
                               "E18 Rugtvedt-Dørdal", "Vestfoldbaneprosjektet"),
                     values = c("springgreen4", "royalblue2",
                                "gold1", "firebrick3")) +
  labs(title = "B", colour = "Excavation project") +
  theme_classic() +
  theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        legend.position = "bottom") +
  guides(colour =  guide_legend(title.position = "top", title.hjust = 0.5))

(overview + sitemap) /
(temp_dist + theme(plot.margin = unit(c(0,135,0,135), "pt")))

A defining characteristic of the Norwegian Mesolithic is that a clear majority of the known sites are located in coastal areas. Furthermore, these coastal sites, including those treated here, appear to predominantly have been located on or close to the contemporary shoreline when they were in use [@breivik2018; @solheim2020]. In south-eastern Norway, this pattern is combined with a continuous regression of the shoreline, following from isostatic rebound [e.g. @sørensen1979; @romundset2018]. The fairly rapid shoreline displacement means that the sites tend not to have retained their strategic or ecologically beneficial shore-bound location for long periods of time [cf. @perreault2019, 47]. Consequently, the shore-bound settlement, combined with the rapid shoreline displacement has resulted in a relatively high degree of spatial separation of cumulative palimpsests, to follow the terminology of @bailey2007, while shoreline displacement combined with shore-bound settlement allows for a relatively good chronological control of these accumulation events. In other parts of the world, a higher degree of spatial distribution means that while the physical separation of material can help delineate discrete events, this typically comes at the cost of loosing temporal resolution, as any stratigraphic relationship between the events is lost [@bailey2007]. However, as the rate of isostatic rebound has varied throughout the Mesolithic in the region, and local topography and bathymetry will have impacted how rapidly a site lost its shore-bound location, this effect is not evenly distributed in time and space. In the earliest part of the Mesolithic, the displacement rate within the study area would have been around as much as 8.8 cm/year, falling to around 0.5 cm/year in the Late Mesolithic [@sørensen2014]. Thus, while relative sea-level change appears to have reduced the degree of mixing that has occurred in the assemblages, this could vary depending on when and where they were in use, in turn potentially reducing the degree to which their composition can be directly compared - a point that is returned to in the discussion of the results.

The data analysed here is based on the classification of the site assemblages done for the original excavation reports, and consists of 48 variables representing different debitage and tool types. While the classification practices for the excavation projects were standardised to an extent, there are some instances where time was allocated to identify additional artefact sub-categories aimed at answering specific research questions. Some categories in the original reports have therefore been combined. This for example pertains to the category narrow-blades, which is defined as blades of width between 8 and 12 mm. In the reports, this was only separated from (macro-)blades (width $\leq$ 12 mm) and micro-blades (width $\leq$ 8 mm) for some of the sites. This was combined with the blade category here. Furthermore, the artefact data is here divided into flint and non-flint materials. Flint does not occur indigenously in Norway, and is only available as nodules that have been transported and deposited by retreating and drifting ice [e.g. @berg-hansen1999]. This means that the distribution and quality of the flint has been impacted by a diverse set of factors relating to climatic and geographical factors such as, but not limited to, topographic variability, shoreline morphology and ocean currents [@eigeland2015, 46]. Consequently, although flint is treated as a unified category here, there is reason to believe that this variability has had consequences for raw-material accessibility and technological choices [@eigeland2015, 45-53]. Furthermore, reducing the number of dimensions by lumping all non-flint materials together also has some issues. A general sentiment is that an increase in the use of non-flint materials signifies an increased adaption to local environments, and possibly a reduction in mobility. However, the various non-flint raw materials can have quite disparate properties, where fine-grained cryptocrystalline materials are often used as a substitute or supplement to flint, while other, coarser materials are usually associated with the production of axes and other macro tools. Given this differentiated use, the retained debitage and tool categories are expected to still reflect the differentiated use associated with different raw-materials. An important benefit of combining all non-flint materials is that this reduces the dependency on whether or not these have been correctly and consistently categorised [cf. @frivoll2017].

@@

The first part of the analysis involves employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports. The purpose of this exercise is to evaluate the degree to which the composition of the assemblages align with patterns that have been suggested by earlier application of more informally driven methods. The expectation was that chronological trends would be the main driver of this part of the analysis - this appears to be largely verfied. While some artefact categories from the original reports are collapsed in an attempt to reduce the amount of subjective discretion that can be expected to have impacted the classification of the individual assemblages, this largely involved an analysis that assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviorally meaningful.

@@

```{r co_data}
# Perform correspondence analysis on the data. 'mean_age' is meant for colouring
# so is excluded here
calithics <- CA(select(lithics, -mean_date, -artefact_count, -flint_count,
                       -reported_start_bp, -reported_end_bp, -askeladden_id),
                graph = FALSE)

@@

The column plot (Figure \@ref(fig:cor)B) reveals that the earliest sites are characterised by the flint artefact categories microburins, projectiles, as well as macro tools and associated debitage. It is also interesting that these sites to larger extent are characerised by core fragments, both in flint and non-flint materials, as opposed to the cores themselves. The non-flint material on the earliest sites appears to be centered around the production of projectiles, as both the projectiles themselves and non-flint blades are important constituents of the assemblages at these sites.

The first dimension, which is pulling some of the later sites towards the right of the plot, is mainly defined by macro tools and associated debitage in non-flint materials that are negatively correlated with more flint dominated assemblages and non-flint projectiles. While the outer end of the first dimensions is dominated by later Mesolithic sites associated with axe production in non-flint materials, the later sites occur along the entire dimension, indicating that while these axe production sites are a feature of the later Mesolithic, there is marked variation among these later sites. Although the sample size is quite strained and the discussion of finer chronological points might not be warranted, the first dimension does appear to be of of less importance for the absolute latest sites, as indicated by their location to the left of the plot. This could indicate that specialised axe production sites disappear towards the end of the Mesolithic, a notion that would be in line with previous research [@eigeland2015, 370].

@@

It has, however, been suggested that access to raw-materials can be a more important decider for assemblage composition than mobility patterns , and that variable workability of these materials will impact factors such as the relative frequency of tools to debitage in the assemblages [e.g. @andrefsky1994; @manninen2014]. In addition to attempts at accounting for this analytically (see below), the limited geographical distribution of the sites should alleviate the issue somewhat, as the overall availability of raw materials could be expected to be more comparable through time than if a wider region was under consideration.

@@

lithics2 <- sites %>% 
  select(-n, -beach_flint_flint_nodule, -hammerstone, -excavation_project,
         -reported_start_bp, -reported_end_bp, -askeladden_id) %>% 
  # The reason for not using .keep below is for the option to 
  #retain artefact_count
  mutate(chip_freq =  sum(chip, chip_other) / artefact_count,
         core_freq = sum(core, core_other) / artefact_count,
         primary_freq = sum(core, core_fragment, blade, 
                     microblade, chip, fragment, flake, flake_other, 
                     fragment_other, chip_other, blade_other,  microblade_other, 
                     core_other, core_fragment_other) / artefact_count,
         secondary_freq =  sum(axe_debitage, axe, chisel, 
                microlith, tanged_point, transverse_point, single_edged_point,
                hognipen_point, projectile_undefined, microburin, scraper, 
                knife, burin, drill, blade_retouch, microblade_retouch, 
                chip_retouch, fragment_retouch, flake_retouch, axe_other,
                axe_debitage_other, chisel_other, club_hatchet_other, 
                grinding_slab_stone, projectile_point_other, knife_other,
                scraper_other, drill_other, microblade_retouch_other, 
                blade_retouch_other, fragment_retouch_other,
                flake_retouch_other) / artefact_count,
         lithic_dens = artefact_count / excavated_m3) %>%
  select(-primary_lithics, -secondary_lithics, -excavated_m3, -artefact_count,
         -flint_count) %>% 
  column_to_rownames("site")

# Repeat but for flint only
flint2 <- sites %>% 
  select(-n, -beach_flint_flint_nodule, -hammerstone, -excavation_project, 
         -reported_start_bp, -reported_end_bp, -askeladden_id) %>% 
  select(!ends_with("other")) %>% 
  mutate(chip_freq =  sum(chip) / artefact_count,
         core_freq = sum(core) / artefact_count,
         primary_freq = sum(core, core_fragment, blade, 
                     microblade, chip, fragment, flake) / artefact_count,
         secondary_freq =  sum(axe_debitage, axe, chisel, 
                microlith, tanged_point, transverse_point, single_edged_point,
                hognipen_point, projectile_undefined, microburin, scraper, 
                knife, burin, drill, blade_retouch, microblade_retouch, 
                chip_retouch, fragment_retouch, flake_retouch) / artefact_count,
         flint_dens = flint_count / excavated_m3) %>%
  select(-primary_flint, -secondary_flint, -excavated_m3, -artefact_count,
         -flint_count) %>% 
  column_to_rownames("site")

@@

```{r wabi, fig.cap = "Relative frequency of secondarily worked lithics plotted against the volumetric density of artefacts (artefact count / excavated m^3^) for A) All lithics, B) Flint. The logarithm is taken to base 10 on all axes.", fig.height = 5, fig.width = 10}
w1 <- ggplot(lithics2, aes(lithic_dens, secondary_freq * 100, 
                     fill = mean_date / 1000)) + 
  geom_point(shape = 21, size = 3) +
  geom_smooth(method = "lm", col = "black", se = FALSE) +
  labs(title = "A",
       x = "Volumetric density of lithics (log10)", 
       y = "Relative frequency of secondarily worked lithics (log10)") +
  scale_fill_gradient2(name = "ka BP",  low="yellow", mid="orange", 
                        high="red", midpoint = centdate) +
  scale_y_log10() +
  scale_x_log10() +
  theme_classic()

w2 <- ggplot(flint2, aes(flint_dens, secondary_freq * 100, 
                     fill = mean_date / 1000)) + 
  geom_point(shape = 21, size = 3) +
  geom_smooth(method = "lm", col = "black", se = FALSE) +
  labs(title = "B", 
       x = "Volumetric density of flint (log10)", 
       y = "Relative frequency of secondarily worked flint (log10)") +
  scale_fill_gradient2(name = "ka BP",  low="yellow", mid="orange", 
                        high="red", midpoint = centdate) +
  scale_y_log10() +
  scale_x_log10() +
  theme_classic()

@@

Figure \@ref(fig:pca-analysis) displays a principle components analysis using variations of the continuous measures for degree of mobility as operationalised by @Bicho2020. While the investigation performed by @Bicho2020 indicates that the relative frequency of chips, cores, and blanks might be more sensitive to mobility patterns, they also found that the relationship between volumetric density of lithics and the frequency of retouched artefacts gave a reasonable estimation of mobility in their data. hese last dimensions are capturing the most variation in the dataset presented here. Thus, if frequency of secondarily worked artefacts is accepted as a proxy for degree of mobility, these findings would consequently be in line with previous research into the Mesolithic of Norway, indicating that earlier sites are associated with higher degree of mobility than sites from later phases.
